---
title: "Do developers trust AI? Adoption, attitudes and pay in the 2025 Stack Overflow Developer Survey"
subtitle: "A complex data analysis in R"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r load, include=FALSE}
source("R/setup.R")
source("R/stats.R")
library(knitr)

d          <- readRDS(PATHS$clean_rds)
models     <- readRDS(file.path(PATHS$models, "models.rds"))
clustering <- readRDS(file.path(PATHS$models, "clustering.rds"))

read_out <- function(name) readr::read_csv(file.path(PATHS$tables, paste0(name, ".csv")),
                                           show_col_types = FALSE)
# Absolute paths, because pandoc resolves image links relative to the .Rmd
# file while knitr chunks run from the project root.
fig <- function(name) knitr::include_graphics(
  normalizePath(file.path("outputs", "figures", paste0(name, ".png"))))
```

# Introduction

AI coding assistants moved from novelty to default tooling in about three
years. This report uses the **Stack Overflow Annual Developer Survey 2025**
(`r format(nrow(d), big.mark = ",")` professional developers after cleaning,
`r n_distinct(d$country, na.rm = TRUE)` countries) to ask who actually adopts
them, who trusts them, and whether any of it shows up in pay.

## Research questions

| # | Question | Method |
|---|----------|--------|
| RQ1 | Does AI tool adoption depend on developer role and experience? | Chi-square, logistic regression |
| RQ2 | Do AI users earn more than non-users? | Welch t-test, Wilcoxon, linear regression |
| RQ3 | Does trust in AI accuracy change with experience? | Chi-square, test for trend |
| RQ4 | Can AI adoption be predicted from professional attributes? | Logistic regression vs random forest |
| RQ5 | Are there distinct developer "personas" in AI attitudes? | PCA, k-means |

## Key findings

```{r headline, include=FALSE}
adopt <- read_out("03_adoption_by_experience")
comp  <- read_out("13_model_comparison")
prof  <- read_out("18_cluster_profiles")
val   <- read_out("19_cluster_validation")
med_user <- median(d$comp_model[d$ai_user == 1], na.rm = TRUE)
med_non  <- median(d$comp_model[d$ai_user == 0], na.rm = TRUE)
```

1. **Adoption is near-universal but shallow in trust.**
   `r round(100 * mean(d$ai_user, na.rm = TRUE), 1)`% use AI tools and
   `r round(100 * mean(d$ai_daily, na.rm = TRUE), 1)`% use them daily, yet only
   `r round(100 * mean(d$ai_trusts, na.rm = TRUE), 1)`% say they trust the
   accuracy of what those tools produce.
2. **Both adoption and trust fall as experience rises.** Adoption drops from
   `r scales::percent(adopt$adoption[1], 0.1)` among developers with 0-2 years
   of experience to `r scales::percent(adopt$adoption[nrow(adopt)], 0.1)` among
   those with 20+ years (chi-square test for trend, p < 0.001).
3. **AI users earn slightly *less*, not more.** Median
   `r scales::dollar(med_user)` versus `r scales::dollar(med_non)`
   (Cohen's d = -0.16). The gap is a small effect and is explained by
   experience and region rather than by AI use itself.
4. **Pay is mostly explained by geography, experience and education.** The
   linear model reaches R² = `r round(broom::glance(models$lm)$r.squared, 3)`;
   AI use contributes almost nothing once those are in the model.
5. **Adoption is hard to predict from professional attributes.** The best model
   reaches AUC `r max(comp$cv_auc, na.rm = TRUE)` (5-fold CV) — better than
   chance, far from decisive.
6. **Trust tracks usage, not seniority in isolation.** An ordinal
   (proportional-odds) model on the full five-point trust scale puts daily
   users at
   `r round(read_out("22_ordinal_trust_model") %>% dplyr::filter(term == "ai_usageDaily") %>% dplyr::pull(odds_ratio), 1)`x
   the odds of higher trust versus never-users, while each additional year of
   experience multiplies those odds by
   `r round(read_out("22_ordinal_trust_model") %>% dplyr::filter(term == "years_code") %>% dplyr::pull(odds_ratio), 3)`.
7. **The complaint is precision, not capability.**
   `r scales::percent(read_out("20_ai_frustrations")$pct[1], 1)` name "AI
   solutions that are almost right, but not quite" as a frustration, and
   `r scales::percent(read_out("20_ai_frustrations")$pct[2], 1)` say debugging
   AI-generated code is more time-consuming than writing it themselves.
8. **Developers split into two clear camps.** k-means finds
   `r clustering$best_k` clusters: an AI-embracing group
   (`r prof$share[1]`% of respondents, `r scales::percent(prof$ai_daily[1], 1)`
   daily users) and a sceptical group (`r prof$share[2]`%,
   `r scales::percent(prof$ai_daily[2], 1)` daily users). The sceptics have a
   higher median salary (`r scales::dollar(val$median_salary[2])` versus
   `r scales::dollar(val$median_salary[1])`) — salary was never used to build
   the clusters.

## Data and licence

Stack Overflow Annual Developer Survey 2025, CC BY-SA 4.0,
<https://survey.stackoverflow.co/>. The raw file holds 172 columns; the
cleaning script keeps `r ncol(d)` after feature engineering and restricts the
sample to people who write code professionally.

# Data preparation

The raw export needs real work before it can be analysed:

* **Types.** `YearsCode` is text because of the answers "Less than 1 year" and
  "More than 50 years"; `Age` is a bucket, converted to a midpoint for
  modelling while the ordered factor is kept for plots.
* **Collapsing.** Education (8 levels), organisation size (9), and country
  (180+) are collapsed to 4, 3 and 6 categories. Rare roles and industries are
  lumped with `fct_lump_min()`. Without this, most tests would be driven by
  categories with a handful of respondents.
* **Salary.** `ConvertedCompYearly` contains values from $1 to tens of millions.
  Values below $1,000 or above the 99th percentile are flagged
  (`comp_outlier`) rather than deleted, and the models use `log(salary)`
  because the distribution is log-normal.
* **Multi-select columns.** `LanguageHaveWorkedWith` and
  `AIModelsHaveWorkedWith` hold semicolon-separated lists; they are split with
  `separate_rows()` for counting and turned into breadth counts
  (`n_languages`, `n_ai_models`) for modelling.
* **Likert scales.** Sentiment and trust become ordered factors *and* numeric
  scores, so they can be both plotted as categories and correlated.

```{r missingness, fig.cap="Missing values by variable"}
fig("01_missingness")
```

Missingness is concentrated in the optional free-text and compensation fields,
which is why every model states the row count it actually fits on.

# Exploratory analysis

```{r overview}
kable(read_out("02_overview"), caption = "Sample at a glance")
```

## Compensation

```{r salary, fig.cap="Compensation distribution"}
fig("02_salary_distribution")
```

```{r salary-region, fig.cap="Compensation by region"}
fig("03_salary_by_region")
```

## AI adoption and attitudes

```{r usage, fig.cap="AI tool usage frequency"}
fig("04_ai_usage")
```

```{r adoption-exp, fig.cap="Adoption by experience"}
fig("05_adoption_by_experience")
```

```{r trust-exp, fig.cap="Trust by experience"}
fig("06_trust_by_experience")
```

```{r sent-trust, fig.cap="Sentiment vs trust"}
fig("07_sentiment_vs_trust")
```

Sentiment and trust are related but far from identical: plenty of developers
are favourable towards AI while still distrusting its output. That gap is the
most interesting single fact in the dataset.

## Technology use

```{r langs, fig.cap="Most used languages"}
fig("09_top_languages")
```

```{r models, fig.cap="Most used AI models"}
fig("10_top_ai_models")
```

```{r frustrations, fig.cap="Reported frustrations with AI tools"}
fig("10b_ai_frustrations")
```

## What developers think survives AI

The survey ends with an open question: *"Looking ahead 3-5 years, what skills
do you believe will remain valuable for developers even as AI tools become
more capable?"* A simple word count over the
`r format(sum(!is.na(d$ai_open) & stringr::str_length(d$ai_open) > 20), big.mark = ",")`
free-text answers shows what people name.

```{r words, fig.cap="Skills expected to remain valuable"}
fig("11_ai_open_words")
```

The answers cluster on judgement rather than syntax: understanding, problem
solving, architecture, design and debugging. That matches the trust results —
developers delegate typing, not deciding.

## Correlations

```{r cor, fig.cap="Correlation heatmap"}
fig("08_correlation_heatmap")
```

# Statistical inference

Every test reports an assumption check, an effect size and a confidence
interval where one exists. Because five families of hypotheses are tested,
p-values are adjusted with the Holm method.

```{r tests}
kable(read_out("09_hypothesis_tests"), caption = "Hypothesis test results")
```

```{r tukey}
kable(read_out("09b_tukey_education"),
      caption = "Tukey HSD: pairwise education comparisons on log(salary)")
```

A note on interpretation: with tens of thousands of respondents, almost
anything reaches statistical significance. The effect sizes matter more than
the p-values, which is why Cramer's V, Cohen's d and eta-squared are reported
alongside every test.

# Regression modelling

## What predicts salary?

```{r lm-coef}
kable(read_out("10_lm_coefficients") %>%
        dplyr::filter(term != "(Intercept)") %>%
        dplyr::arrange(dplyr::desc(abs(estimate))) %>%
        head(15),
      caption = "Linear model on log(salary): largest 15 effects. `pct_effect` is the percentage change in salary.")
```

```{r lm-r2}
gl <- broom::glance(models$lm)
cat("R-squared:", round(gl$r.squared, 3),
    "| Adjusted:", round(gl$adj.r.squared, 3),
    "| Observations:", gl$nobs)
```

```{r lm-effects, fig.cap="Largest salary effects"}
fig("14_lm_effects")
```

### Diagnostics

```{r lm-resid, fig.cap="Residuals vs fitted"}
fig("12_lm_residuals")
```

```{r lm-qq, fig.cap="Q-Q plot"}
fig("13_lm_qq")
```

```{r vif}
kable(read_out("11_lm_vif"),
      caption = "Variance inflation factors. GVIF^(1/(2*Df)) above ~2.2 signals collinearity.")
```

Collinearity is not a problem here (every scaled GVIF is close to 1), but
**constant variance is**. The residual plot fans out, and a Breusch-Pagan test
confirms it: BP = `r round(models$bp_test$statistic, 1)`,
df = `r models$bp_test$parameter`, p `r fmt_p(models$bp_test$p.value)`.
Ordinary standard errors are therefore wrong — the coefficients stay unbiased,
but their precision is overstated.

```{r robust}
rob <- read_out("21_lm_robust_se")
kable(rob %>%
        dplyr::arrange(dplyr::desc(se_ratio)) %>%
        head(10) %>%
        dplyr::select(term, estimate, se_classical, se_robust, se_ratio,
                      p_classical, p_robust, flips_at_5pct),
      caption = "HC3 sandwich standard errors versus the classical ones, ten largest ratios.")
```

Robust errors are on average `r round(mean(rob$se_ratio), 3)`x the classical
ones and `r sum(rob$flips_at_5pct)` coefficients change their verdict at the 5%
level. Every conclusion below is stated against the robust errors.

### Does dropping the non-reporters bias this?

Only `r round(100 * mean(!is.na(d$comp_yearly)), 1)`% of respondents answer the
compensation question, and the test in H6 shows the response rate depends on
region. Refitting the model weighted by the inverse probability of reporting —
so under-reporting groups stand in for their missing peers — is the direct
check.

```{r ipw}
ipw <- read_out("25_ipw_sensitivity")
kable(head(ipw, 8),
      caption = "Complete-case versus inverse-probability-weighted coefficients, largest shifts first.")
```

The median absolute shift is
`r round(median(abs(ipw$difference), na.rm = TRUE), 4)` log points — under half
a percent of salary. The complete-case results survive the objection.

## Who uses AI tools?

```{r logit}
kable(read_out("12_logit_odds_ratios") %>%
        dplyr::filter(term != "(Intercept)") %>%
        dplyr::arrange(dplyr::desc(abs(log(odds_ratio)))) %>%
        head(15),
      caption = "Logistic regression, odds ratios (>1 means more likely to use AI tools)")
```

```{r comparison}
kable(read_out("13_model_comparison"),
      caption = "Held-out performance. `cv_auc` is the mean AUC over 5 folds.")
```

Four in five respondents use AI tools, so a model that predicts "user" for
everyone already scores
`r scales::percent(read_out("13_model_comparison")$accuracy[3], 0.1)` accuracy
while being useless. The decision threshold is therefore tuned on the training
set (Youden's J) rather than left at 0.5, which trades raw accuracy for a model
that actually separates the two groups — and makes AUC, not accuracy, the
metric worth reading in that table.

```{r roc, fig.cap="ROC curves"}
fig("15_roc_curves")
```

```{r importance, fig.cap="Random forest importance"}
fig("16_rf_importance")
```

The two models land close together, which is itself the finding: the
relationship between professional attributes and AI adoption is mostly
additive, so the random forest's extra flexibility buys little. Both beat the
majority-class baseline, but neither is a strong classifier — adoption is
driven by things the survey does not measure (team policy, employer tooling,
personal taste).

## What predicts trust in AI accuracy?

Trust is recorded on a five-point ordered scale. The chi-square tests above
collapse it to a binary, which throws away the ordering; treating it as a
number would assume the gap between "somewhat distrust" and "neither" equals
the gap between "somewhat trust" and "highly trust". A proportional-odds model
(`MASS::polr`, n = `r format(models$ordinal_n, big.mark = ",")`) needs neither
assumption.

```{r ordinal}
ord <- read_out("22_ordinal_trust_model")
kable(ord %>%
        dplyr::filter(kind == "predictor") %>%
        dplyr::arrange(dplyr::desc(abs(estimate))) %>%
        head(14) %>%
        dplyr::select(term, odds_ratio, conf_low, conf_high, p_value),
      caption = "Ordinal logistic regression on trust. Odds ratio > 1 means higher trust.")
```

```{r ordinal-fig, fig.cap="Predictors of trust in AI accuracy"}
fig("22_ordinal_trust_effects")
```

Usage dominates everything else: daily users have
`r round(ord$odds_ratio[ord$term == "ai_usageDaily"], 1)`x the odds of a higher
trust rating than people who never use AI tools. Experience pushes the other
way — each additional year of coding multiplies the odds by
`r round(ord$odds_ratio[ord$term == "years_code"], 3)`, which compounds to
roughly `r round(100 * (1 - ord$odds_ratio[ord$term == "years_code"]^10), 0)`%
lower odds over a decade. The direction of causation is not identified here:
trusting the tools and using them daily reinforce each other.

```{r po-check}
po <- read_out("23_proportional_odds_check")
kable(head(po, 8),
      caption = "Proportional-odds check: coefficients refitted at each of the four cut-points.")
```

The assumption mostly holds: `r sum(po$unstable)` of `r nrow(po)` coefficients
vary more across cut-points than their own size, and they are sparse
role categories (retired developers, senior executives) rather than the
variables the conclusions rest on.

# Developer personas

PCA and k-means on the AI-attitude and experience variables, with *k* chosen
by average silhouette width.

```{r scree, fig.cap="PCA scree plot"}
fig("17_pca_scree")
```

```{r loadings}
kable(read_out("16_pca_loadings"), caption = "PCA loadings, first three components")
```

```{r elbow, fig.cap="Elbow plot"}
fig("18_kmeans_elbow")
```

```{r sil, fig.cap="Silhouette widths"}
fig("19_kmeans_silhouette")
```

```{r centres, fig.cap="Cluster centres"}
fig("20_cluster_centres")
```

```{r cluster-pca, fig.cap="Clusters in PC space"}
fig("21_cluster_pca")
```

```{r profiles}
kable(read_out("18_cluster_profiles"), caption = "Cluster profiles, original units")
```

### How solid is this split?

Two checks, and they disagree in an informative way.

```{r stability-tbl}
stab <- read_out("24_cluster_stability")
kable(stab, caption = "Silhouette width and resampling stability")
```

```{r stability-fig, fig.cap="Cluster stability under resampling"}
fig("23_cluster_stability")
```

The **silhouette width is `r stab$silhouette`** — below the 0.25 that Kaufman
and Rousseeuw call the threshold for "substantial structure". Taken alone, that
says the two groups are not cleanly separated blobs.

The **adjusted Rand index is `r stab$mean_ari`** across `r stab$runs` runs on
80% subsamples (minimum `r stab$min_ari`), meaning a re-clustered subsample
reproduces almost exactly the same partition.

Read together: developers lie on a continuum of AI attitude with no gap in the
middle, but the line k-means draws through that continuum is in a stable,
reproducible place. The split is a real description, not a separate-species
claim — and the validation below shows it predicts variables it never saw.

```{r validation}
kable(read_out("19_cluster_validation"),
      caption = "Validation on variables that were not used to build the clusters")
```

The validation table is the point of this section: salary, remote work and
organisation size were deliberately excluded from the clustering features, so
differences across clusters on those variables are evidence that the personas
capture something real rather than an artefact of the algorithm.

# Limitations

* **Self-selected sample.** Respondents are Stack Overflow visitors who chose
  to answer; they over-represent North America and Europe, so nothing here
  generalises to the whole developer population.
* **Cross-sectional.** Every association is correlational. Higher earners
  using AI more does not mean AI raised anyone's salary; both are downstream of
  employer, seniority and country.
* **Converted compensation.** `ConvertedCompYearly` depends on exchange rates
  and self-reported figures, and the currency conversion date is not exposed.
* **Collapsed categories.** Grouping 180 countries into 6 regions and 8
  education levels into 4 loses real variation. It trades precision for models
  that fit and tests that have adequate cell counts.
* **Large n.** Statistical significance is nearly guaranteed at this sample
  size; effect sizes are the honest summary.
* **Missingness is addressed, not eliminated.** The inverse-probability
  weighting corrects for reporting patterns that the *observed* predictors
  explain. If people hide their salary for reasons unrelated to region, role
  or experience — being unusually badly paid, say — that bias remains.
* **Prediction is weak by design of the data.** AUC 0.66 for adoption is an
  honest ceiling given what the survey asks: employer policy, team norms and
  tooling mandates are the obvious missing predictors.

# Reproducing this report

```bash
Rscript scripts/00_setup.R     # install packages, download data (~141 MB)
Rscript scripts/run_all.R      # clean, analyse, model, render this report
```

All randomness is seeded (`SEED = 42` in `R/setup.R`), so reruns reproduce the
same splits, folds and clusters.
